Fishing for Spam

### 2024-07-18 code/analysis-code/CopyOfstatistical-analysis.qmd
Further exploration, prepping data for modeling, model tuning and multiple feature set iterations

```{r}
# load needed packages. make sure they are installed.
pacman::p_load(here, #for data loading/saving
               tidyverse, # for data manipulation and plotting
               skimr, # for data exploration and summary
               ggthemes, # for plot theming
               tidytext, # for text analysis
               quanteda, # for text analysis
               caret, # for model training
               e1071, # for model training
               randomForest, # for model training
               irlba, # for SVD
               doSNOW, # for parallel processing
               doParallel, # for parallel processing
               parallel) # for parallel processing

@@

```{r}
# create smaller partition for the sake of being able to run this shit on my PC
subsample <- createDataPartition(mydata$label,
                                 times = 1,
                                 p = .1,
                                 list = FALSE)

@@

# create index of partition
inTrain <- createDataPartition(sub.data$label, 
                               times = 1, 
                               p = 0.7,
                               list = FALSE)

@@

```{r}
# create CV folds and model controls
folds <- createMultiFolds(train$label,
                          k = 10,
                          times = 3)
ctrl <- trainControl(method = "repeatedcv",
                     number = 10,
                     repeats = 3,
                     index = folds)
gridSVM <- expand.grid(C = seq(.1, 1, by = .1),
                       sigma = c(0.01, 0.1, 0.5, 1))

@@

# project singular value decomp onto test split
projSVD <- function(trainirlba, testtfidf) {
  sigma.inverse <- 1 / trainirlba$d
  u.transpose <- t(trainirlba$u)
  projection <- t(sigma.inverse * u.transpose %*% t(testtfidf))
  projection <- as.data.frame(projection)
  names(projection) <- paste0("X", seq(1,300))
  return(projection)
}

# tune Random Forest model
customRF <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1, type = "SOCK") # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "rf",
                 trControl = ctrl,
                 tuneLength = 7)
  stopCluster(clus)
  return(model)
}

# tune an SVM model
customSVM <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1, type = "SOCK") # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "svmRadial",
                 trControl = ctrl,
                 tuneLength = 7,
                 tuneGrid = gridSVM)
  stopCluster(clus)
  return(model)
}
```

@@

@@

Finally, we're left with a matrix containing 40,399 features. While many classification models can ultimately accommodate such intense dimensionality, the computational resource requirements are quite high and would therefore make it unfeasible to run these models with the limited memory and processing power available on consumer-grade hardware. Additionally, the sparsity of our matrix (\>99.9%) suggests that many of the features will not be particularly predictive of either classification on their own.

@@

saveRDS(train.sub.irlba, file = here("results", "large-files", "train-sub-irlba.rds"))
```

@@

####### Subject Token Only #######
# create revised training data frame with subject token SVD
train.sub.svd <- data.frame(label = train$label,
                            train.sub.irlba$v)
saveRDS(train.sub.svd, file = here("results", "large-files", "train-sub-svd.rds"))

# create revised training data frame with subject token SVD and email length
train.sub.svd.length <- data.frame(label = train$label,
                                   length = train$bodylength,
                                   train.sub.irlba$v)
saveRDS(train.sub.svd.length, file = here("results", "large-files", "train-sub-svd-length.rds"))

# create revised training data frame 
# with subject token SVD, email length, and sender domain
train.sub.svd.length.sender <- data.frame(label = train$label,
                                          length = train$bodylength,
                                          senderdomain = train$senderdomain,
                                          train.sub.irlba$v)
saveRDS(train.sub.svd.length.sender, file = here("results", "large-files", "train-sub-svd-length-sender.rds"))

####### All Tokens #######
# create revised training data frame with all tokens SVD
train.all.svd <- data.frame(label = train$label,
                            train.body.irlba$v,
                            train.sub.irlba$v)
saveRDS(train.all.svd, file = here("results", "large-files", "train-all-svd.rds"))

# create revised training data frame with all tokens SVD and email length
train.all.svd.length <- data.frame(label = train$label,
                                   length = train$bodylength,
                                   train.body.irlba$v,
                                   train.sub.irlba$v)
saveRDS(train.all.svd.length, file = here("results", "large-files", "train-all-svd-length.rds"))

# create revised training data frame 
# with all tokens SVD, email length, and sender domain
train.all.svd.length.sender <- data.frame(label = train$label,
                                          length = train$bodylength,
                                          senderdomain = train$senderdomain,
                                          train.body.irlba$v,
                                          train.sub.irlba$v)
saveRDS(train.all.svd.length.sender, file = here("results", "large-files", "train-all-svd-length-sender.rds"))
```

@@

@@

@@

```{r}
# create document-feature matrix
test.body.tokens.dfm <- dfm(test.body.tokens,
                             tolower = FALSE)
```

```{r}
# project the feature set from the train split onto the test split
test.body.tokens.dfm <- dfm_match(test.body.tokens.dfm, features = featnames(train.body.tokens.dfm))
```

```{r}
# convert dfm to matrix
test.body.tokens.matrix <- as.matrix(test.body.tokens.dfm)
```

```{r}
test.body.tokens.tf <- apply(test.body.tokens.matrix, 1, term.freq)
```

```{r}
test.body.tokens.tfidf <- apply(test.body.tokens.tf, 2, tf.idf, idf = train.body.tokens.idf)
```

```{r}
# transpose and fix missing values
test.body.tokens.tfidf <- t(test.body.tokens.tfidf)
test.body.tokens.tfidf[is.na(test.body.tokens.tfidf)] <- 0.0
```

@@

@@

# Generate Predictions

```{r}
# predict test data with model
#rf.body.preds <- predict(rf.body, test.body.svd)
#rf.body.length.preds <- predict(rf.body.length, test.body.svd.length)
#rf.body.length.sender.preds <- predict(rf.body.length.sender, test.body.svd.length.sender)
```

The `echo: false` option disables the printing of code (only output is displayed).

>>>>

```{r}
# load needed packages. make sure they are installed.
pacman::p_load(here, #for data loading/saving
               tidyverse, # for data manipulation and plotting
               tidytext, # for text analysis
               quanteda, # for text analysis
               caret, # for model training
               e1071, # for model training
               randomForest, # for model training
               xgboost, # for model training
               irlba, # for SVD
               doParallel, # for parallel processing
               parallel) # for parallel processing

@@

However, before we can proceed with either step we first need to ensure that the size of our data is manageable for the scope of our machine. Modeling a very large corpus of unstructured text data incurs high computational costs due to the increasing dimensionality of the data as we begin to structure it. As a result, the amount of data that can be modeled will vary significantly based on the computational resources available to you.

At this time we've made the decision to examine a 10% sample of our overall data set for the purpose of modeling on our machine, but the proportion parameter for the subsample partition can be tweaked according to your available resources.

```{r}
# create smaller partition for the sake of being able
# to run on consumer grade hardware 
subsample <- createDataPartition(mydata$label,
                                 times = 1,
                                 p = .1,
                                 list = FALSE)

@@

# create index of train partition
inTrain <- createDataPartition(sub.data$label, 
                               times = 1, 
                               p = 0.7,
                               list = FALSE)

@@

```{r}
# create CV folds and model controls
folds <- createMultiFolds(train$label,
                          k = 10,
                          times = 3)
ctrl <- trainControl(method = "repeatedcv",
                     number = 10,
                     repeats = 3,
                     index = folds,
                     allowParallel = TRUE)
gridSVM <- expand.grid(C = seq(.1, 1, by = .1),
                       sigma = c(0.01, 0.1, 0.5, 1))
gridXG <- expand.grid(nrounds = c(100, 200),
                      max_depth = c(3, 6),
                      eta = c(0.1, 0.3),
                      gamma = c(0, 0.1),
                      colsample_bytree = c(0.8),
                      min_child_weight = c(1, 3),
                      subsample = c(0.8))

@@

# project singular value decomp onto test split
projSVD <- function(type, trainirlba, testtfidf) {

  sigma.inverse <- 1 / trainirlba$d
  u.transpose <- t(trainirlba$u)
  projection <- t(sigma.inverse * u.transpose %*% t(testtfidf))
  projection <- as.data.frame(projection)
  if (type == "bod") {
    names(projection) <- paste0("X", seq(1,300))
  } else {
    names(projection) <- paste0("Z", seq(1,300))
  }
  return(projection)
}

# tune Random Forest model
customRF <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1) # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "rf",
                 trControl = ctrl,
                 tuneLength = 7,
                 verbose = FALSE)
  stopCluster(clus)
  return(model)
}

# tune an SVM model
customSVM <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1) # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "svmRadial",
                 trControl = ctrl,
                 tuneLength = 7,
                 tuneGrid = gridSVM,
                 verbose = FALSE)
  stopCluster(clus)
  return(model)
}

# tune an XGBoost model
customXG <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1) # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "xgbTree",
                 trControl = ctrl,
                 tuneGrid = gridXG,
                 verbose = FALSE)
  stopCluster(clus)
  return(model)
}
```

@@

Finally, we're left with a matrix containing 40,399 features. While many classification models can ultimately accommodate such intense dimensionality, the computational resource requirements are quite high and would therefore make it unfeasible to run these models with the limited memory and processing power available on consumer-grade hardware. Additionally, the sparsity of our matrix (>99.9%) suggests that many of the features will not be particularly predictive of either classification on their own.

@@

@@

saveRDS(train.sub.irlba, file = here("results", "large-files", "train-sub-irlba.rds"))

sub.svd <- as.data.frame(train.sub.irlba$v)
sub.svd <- sub.svd %>%
  rename_all(~ paste0("Z", seq(1, length(names(sub.svd)))))
```

@@

####### Subject Token Only #######
# create revised training data frame with subject token SVD
train.sub.svd <- data.frame(label = train$label,
                            sub.svd)
saveRDS(train.sub.svd, file = here("results", "large-files", "train-sub-svd.rds"))

# create revised training data frame with subject token SVD and email length
train.sub.svd.length <- data.frame(label = train$label,
                                   length = train$bodylength,
                                   sub.svd)
saveRDS(train.sub.svd.length, file = here("results", "large-files", "train-sub-svd-length.rds"))

# create revised training data frame 
# with subject token SVD, email length, and sender domain
train.sub.svd.length.sender <- data.frame(label = train$label,
                                          length = train$bodylength,
                                          senderdomain = train$senderdomain,
                                          sub.svd)
saveRDS(train.sub.svd.length.sender, file = here("results", "large-files", "train-sub-svd-length-sender.rds"))

####### All Tokens #######
# create revised training data frame with all tokens SVD
train.all.svd <- data.frame(label = train$label,
                            train.body.irlba$v,
                            sub.svd)
saveRDS(train.all.svd, file = here("results", "large-files", "train-all-svd.rds"))

# create revised training data frame with all tokens SVD and email length
train.all.svd.length <- data.frame(label = train$label,
                                   length = train$bodylength,
                                   train.body.irlba$v,
                                   sub.svd)
saveRDS(train.all.svd.length, file = here("results", "large-files", "train-all-svd-length.rds"))

# create revised training data frame 
# with all tokens SVD, email length, and sender domain
train.all.svd.length.sender <- data.frame(label = train$label,
                                          length = train$bodylength,
                                          senderdomain = train$senderdomain,
                                          train.body.irlba$v,
                                          sub.svd)
saveRDS(train.all.svd.length.sender, file = here("results", "large-files", "train-all-svd-length-sender.rds"))
```

@@

## XGBoost Iterations

```{r}
###### Body Tokens Only ######
# tune an XGBoost model with body token SVD
xg.body.svd <- customXG(train.body.svd)

saveRDS(xg.body.svd, file = here("results", "large-files", "xg-body-svd.rds"))

# tune an XGBoost model with body token SVD and email length
xg.body.svd.length <- customXG(train.body.svd.length)

saveRDS(xg.body.svd.length, file = here("results", "large-files", "xg-body-svd-length.rds"))

### can't be run in a reasonable timeframe
### too computationally intensive for a MacBook
# tune an XGBoost model with body token SVD, email length, and sender domain
#xg.body.svd.length.sender <- customXG(train.body.svd.length.sender)

#saveRDS(xg.body.svd.length.sender, file = here("results", "large-files", "xg-body-svd-length-sender.rds"))

###### Subject Tokens Only ######
# tune an XGBoost model with subject token SVD
xg.sub.svd <- customXG(train.sub.svd)

saveRDS(xg.sub.svd, file = here("results", "large-files", "xg-sub-svd.rds"))

# tune an XGBoost model with subject token SVD and email length
xg.sub.svd.length <- customXG(train.sub.svd.length)

saveRDS(xg.sub.svd.length, file = here("results", "large-files", "xg-sub-svd-length.rds"))

### can't be run in a reasonable timeframe
### too computationally intensive for a MacBook
# tune an XGBoost model with subject token SVD, email length, and sender domain
#xg.sub.svd.length.sender <- customXG(train.sub.svd.length.sender)

#saveRDS(xg.sub.svd.length.sender, file = here("results", "large-files", "xg-sub-svd-length-sender.rds"))

###### All Tokens ######
# tune an XGBoost model with all token SVD
xg.all.svd <- customXG(train.all.svd)

saveRDS(xg.all.svd, file = here("results", "large-files", "xg-all-svd.rds"))

# tune an XGBoost model with all token SVD and email length
xg.all.svd.length <- customXG(train.all.svd.length)

saveRDS(xg.all.svd.length, file = here("results", "large-files", "xg-all-svd-length.rds"))

### can't be run in a reasonable timeframe
### too computationally intensive for a MacBook
# tune an XGBoost model with all token SVD, email length, and sender domain
#xg.all.svd.length.sender <- customXG(train.all.svd.length.sender)

#saveRDS(xg.all.svd.length.sender, file = here("results", "large-files", "xg-all-svd-length-sender.rds"))
```

@@

Having now run and saved our respective iterations of various models and features we can now proceed with preparing the test data so that it exists in the same features space with the same data transformations. The first step of which is to follow tokenization utilizing the same exact methods as we did for the training data.

@@

We begin to deviate from our processing workflow utilized in the training section in order to ensure that the data is mapped into the same feature space as the training data used for modeling. In order to ensure this we'll utilize the `dfm_match()` function to project the same token space from the training data onto our test data.

```{r}
# create document-feature matrix
test.body.tokens.dfm <- dfm(test.body.tokens,
                             tolower = FALSE)

# project the feature set from the train split onto the test split
test.body.tokens.dfm <- dfm_match(test.body.tokens.dfm, features = featnames(train.body.tokens.dfm))

# convert dfm to matrix
test.body.tokens.matrix <- as.matrix(test.body.tokens.dfm)
```

Similarly we need to ensure the inverse document frequency values used to discount overrepresented terms are the same as those utilized in our training data when tranforming to TF-IDFs.

```{r}
test.body.tokens.tf <- apply(test.body.tokens.matrix, 1, term.freq)

test.body.tokens.tfidf <- apply(test.body.tokens.tf, 2, tf.idf, idf = train.body.tokens.idf)

# transpose and fix missing values
test.body.tokens.tfidf <- t(test.body.tokens.tfidf)
test.body.tokens.tfidf[is.na(test.body.tokens.tfidf)] <- 0.0
```

Finally we'll be using the same SVD process as was used on the training data. We'll project the data using our custom `projSVD()` function defined in the Controls and Functions section above.

@@

Having now completed the transformations necessary to map the test data into the same space as the training data used for modeling we can proceed with making predictions and assessing model performance. The latter of which will be continued in a second statistical analysis `.QMD` file using the pertinent, saved `.RDS` files from this document.

@@

```{r}
xg.body.svd.preds <- predict(xg.body.svd, test.body.svd)
confusionMatrix(xg.body.svd.preds, test.body.svd$label)
xg.all.svd.length.preds <- predict(xg.all.svd.length, test.all.svd.length)
confusionMatrix(xg.all.svd.length.preds, test.body.svd$label)
```

### 2024-07-18 code/analysis-code/statistical-analysis-continued.qmd
Further exploration, prepping data for modeling, model tuning and multiple feature set iterations

# tune Random Forest model
customRF <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1, type = "SOCK") # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "rf",
                 trControl = ctrl,
                 tuneLength = 7)
  stopCluster(clus)
  return(model)
}

# tune an SVM model
customSVM <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1, type = "SOCK") # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "svmRadial",
                 trControl = ctrl,
                 tuneLength = 7,
                 tuneGrid = gridSVM)
  stopCluster(clus)
  return(model)
}
```

@@

# Generate Predictions

```{r}
# predict test data with model
#rf.body.preds <- predict(rf.body, test.body.svd)
#rf.body.length.preds <- predict(rf.body.length, test.body.svd.length)
#rf.body.length.sender.preds <- predict(rf.body.length.sender, test.body.svd.length.sender)
```

>>>>

# tune Random Forest model
customRF <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1) # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "rf",
                 trControl = ctrl,
                 tuneLength = 7)
  stopCluster(clus)
  return(model)
}

# tune an SVM model
customSVM <- function(df) {
  cores <- detectCores()
  clus <- makeCluster(cores[1] - 1) # determine the optimal number
                                                   # of cores for parallel processing
  registerDoParallel(clus)
  model <- train(label ~ ., 
                 data = df,
                 method = "svmRadial",
                 trControl = ctrl,
                 tuneLength = 7,
                 tuneGrid = gridSVM)
  stopCluster(clus)
  return(model)
}
```

@@

### 2024-07-18 products/manuscript/Manuscript.qmd
Further exploration, prepping data for modeling, model tuning and multiple feature set iterations

---
title: "Fishing for Spam"
subtitle: "Part 1: Project Proposal"
author: Sean O'Sullivan
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

@@

The data set(s) that I'd liked to examine are a handful of annotated data sets pertaining to fraudulent (spam) emails. They come from a handful of sources, te largest of the bunch is from the Conference on Email and Anti-Spam. That data set and a pair of others found on [Kaggle](https://www.kaggle.com/datasets/naserabdullahalam/phishing-email-dataset/data?select=CEAS_08.csv) contain text information from the subjects and bodies of the emails as well as the senders, receivers, timestamps, and whether or not URLs are contained within.

@@

I'd like to explore the features that are prevalent in the identification of spam emails. Utilizing text analysis in addition to the other email attributes supplied my end goal would be to model a classifier that has a high rate of efficacy in identifying spam emails for automated detection as used by some popular email providers and large internal IT departments across the corporate landscape.

@@

![](../../results/figures/bodylength-distribution.png){.lightbox} First examining the distribution of email body length between spam and legitimate emails we observe a tendency for spam emails to be shorter in length. However, interestingly the distribution of spam emails is bi-modal -- with a second peak at around the 4,000 character count -- which may indicate that spam vendors are aware of the identification potential of email length and have begun targeting a longer length to avoid notice.

![](../../results/figures/subjectlength-distribution.png){.lightbox} Similarly, when we examine subject line length we observe the same tendency for spam emails to trend shorter. Subject line does not share the same distinct bi-modal shape for spam emails that the bodies do, but subject line length distribution between both spam and legitimate emails overlap in large part which may reduce it's efficacy for classification modeling.

::: {layout-ncol="2"}
![](../../results/figures/spam-by-domain.png)

![](../../results/figures/legit-by-domain.png)
:::

@@

>>>>

---
title: "Fishing for Spam"
author: Sean O'Sullivan
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

```{r, echo=FALSE, message=FALSE}
# load needed packages. make sure they are installed.
pacman::p_load(here, #for data loading/saving
               tidyverse, # for data manipulation and plotting
               skimr, # for data exploration and summary
               ggthemes) # for plot theming
```

@@

The data set(s) that I'd liked to examine are a handful of annotated data sets pertaining to fraudulent (spam) emails. They come from a handful of sources, the largest of the bunch is from the Conference on Email and Anti-Spam. That data set and a pair of others can be found on [Kaggle](https://www.kaggle.com/datasets/naserabdullahalam/phishing-email-dataset/data?select=CEAS_08.csv) contain text information from the subjects and bodies of the emails as well as the senders, receivers, timestamps, and whether or not URLs are contained within.

@@

I'd like to explore the features that are prevalent in the identification of spam emails. Utilizing text analysis, in addition to the other email attributes supplied and engineered, my end goal would be to model a classifier that has a high rate of efficacy in identifying spam emails for use of automated detection by some popular email providers and large internal IT departments across the corporate landscape.

@@

First examining the distribution of email body length between spam and legitimate emails we observe a tendency for spam emails to be shorter in length. However, interestingly the distribution of spam emails is bi-modal -- with a second peak at around the 4,000 character count -- which may indicate that spam vendors are aware of the identification potential of email length and have begun targeting a longer length to avoid notice. ![](../../results/figures/bodylength-distribution.png){.lightbox}

Similarly, when we examine subject line length we observe the same tendency for spam emails to trend shorter. Subject line does not share the same distinct bi-modal shape for spam emails that the bodies do, but subject line length distribution between both spam and legitimate emails overlap in large part which may reduce it's efficacy for classification modeling. ![](../../results/figures/subjectlength-distribution.png){.lightbox}

When plotting body tokens we can see that 'http', which precedes most links within each email, is the most common token for each label. Considering this token should approximate not just the presence of a URL, but also the quantity of URLs within each individual email, it will be interesting to see if including the supplied `URLs` boolean will be of value or not in our end model. Given the below, we suspect that the overlap will result in unnecessary noise and dimensionality to our final model. Additionally, considering the above bar chart suggests link presence and quantity is very similarly distributed across both labels, it may wind up improving model performance altogether to exclude 'http' or URL features from the final model.

![](../../results/figures/popular-body-tokens.png)

@@

![](../../results/figures/popular-domains.png)
